openTSNE: a modular \proglang{Python} library for t-SNE dimensionality reduction and embedding

### 2021-05-13 paper/main.tex
minor changes

%% - \Abstract{} almost as usual
\Abstract{
One of the most popular techniques for constructing visualizations of large,
high-dimensional data sets is $t$-distributed stochastic neighbor embedding
(t-SNE). Recently, several extensions have been proposed to address
issues of scalability and the quality of the resulting visualizations. We
introduce \opentsne, a modular Python library that implements the core t-SNE
algorithm and its many extensions. The library is fast and scales up to data
sets containing millions of data points. \opentsne\ is distributed under the
BSD-3-Clause License and its source code is publicly available at
\url{https://github.com/pavlin-policar/openTSNE}.
}

@@

The ever-growing volumes of high-dimensional data sets in machine learning calls
for efficient dimensionality reduction techniques and their implementations,
capable of producing informative data visualizations. Popular approaches include
principal component analysis, multidimensional scaling, $t$-distributed stochastic
neighbor embedding (t-SNE)~\citep{maaten2008visualizing}, and uniform manifold
approximation and projections (UMAP)~\citep{2018arXivUMAP}. Among these, t-SNE
has received much attention as it can address high volumes of data and reveal
the underlying data structure. For instance, t-SNE is widely used in the
bioinformatics community in areas such as single-cell
transcriptomics~\citep{macosko2015highly,cao2019single,tasic2018shared}, human
genetics~\citep{hirata2019genetic}, metagenomic
assembly~\citep{beaulaurier2018metagenomic}, the spatial organization of
microbial communities~\citep{sheth2019spatial}, and
metabolomics~\citep{tkachev2019differences}. Reports on single-cell gene
expression data, our running example, often start with an overview of the cell
landscape, where t-SNE is used to embed high-dimensional expression profiles
into a two-dimensional space. Figs.~\ref{fig:macosko}.a and \ref{fig:macosko}.b
show two such embeddings.

\begin{figure*}[ht]
  \includegraphics[width=\textwidth]{macosko2015}
  \caption{\label{fig:macosko}
  We use \opentsne\ to generate three t-SNE embeddings and demonstrate recent
  theoretical advances. The data in \textbf{(a)} and \textbf{(b)} represent
  44,808 single-cell gene-expression profiles of mouse retinal cells from 
  \citet{macosko2015highly}. The data in \textbf{(c)} additionally contains
  27,499 expression profiles from mouse retinal cells from
  \citet{shekhar2016comprehensive}. \textbf{(a)} We construct a t-SNE embedding
  following the parameter choices from the original publication by
  \citet{maaten2008visualizing}. The visualization shows no preservation of the
  global organization of clusters, resulting from random initialization and an
  affinity model focused on preserving local neighborhoods. \textbf{(b)} A
  modern t-SNE embedding, utilizing the latest theoretical advances and
  practical recommendations constructed using a multi-scale affinity model,
  preserving both short-range and long-range interactions between data points
  and initialized so that the global layout is as meaningful as possible. Unlike
  in \textbf{(a)}, the green and blue clusters representing different sub-types
  of amacrine and bipolar cells are now localized to the same regions of the
  space, indicating a higher level of similarity than to other cell types. The
  embedding in \textbf{(c)} shows how existing t-SNE reference atlases can be
  used to place new samples into existing embeddings. The positions of new data
  points correspond to cell types from the reference atlas.
}
\end{figure*}

Despite its utility, t-SNE has often been criticized for its limited
scalability, lack of global organization -- t-SNE identifies well-defined
clusters that may be arbitrarily scattered throughout the low-dimensional space
-- and the absence of theoretically-founded methods to map new data into
existing embeddings~\citep{ding2018interpretable,becht2019dimensionality}. Most
of these shortcomings have recently been addressed. \citet{linderman2019fast}
developed FIt-SNE, an efficient approximation scheme that massively improves
the scalability of t-SNE, achieving linear time complexity in the number of
samples. \citet{kobak2019art} proposed several techniques to improve global
cluster coherence, including estimating similarities with a mixture of Gaussian
kernels. In our previous work, we introduced a principled approach for embedding
new samples into existing visualizations~\citep{policar2019embedding}.

However, easy access to efficient implementations almost universally correlates
with the widespread adoption of novel computational techniques. Despite the many
theoretical advancements, most popular t-SNE libraries have been slow to
incorporate them into their implementations. In this paper, we present \opentsne,
our open-source Python implementation of t-SNE and its many recently proposed
extensions. \opentsne\ is easy to install and includes precompiled binaries
available through popular Python package managers. It provides a familiar API,
making it suitable as a drop-in replacement for existing t-SNE implementations.

@@

In standard t-SNE, distances between data points are converted to similarities
through the use of Gaussian kernels. The perplexity of these kernels governs the
sizes of the local neighborhoods which are to be preserved. One easy way to
improve the global consistency of resulting embeddings is to increase the sizes
of these neighborhoods. This can be achieved by increasing the perplexity
parameter. However, this often comes at the price of local structure.
\citet{kobak2019art} propose to instead use mixtures of Gaussians to better
preserve short-range, as well as long-range distances and, thus, achieve a
better trade-off between local and global structure.

Embeddings produced by standard t-SNE often make use of all available space and
separates clusters by only thin boundaries. When working with large data sets,
this often obscures the global relationships between clusters as all neighboring
groups appear equidistant from one another.  Other dimensionality reduction
methods, such as UMAP and ForceAtlas2, produce embeddings where clusters appear
more compact and the white space separating the clusters may be interpreted as a
loose measure of distance. Recently, \citet{bohm2020unifying} showed that the
exaggeration factor $\rho$ could be used to produce layouts more similar to UMAP
and ForceAtlas2. By incorporating exaggeration into later phases of the
optimization, t-SNE introduces more white space between clusters which better
reflects the global relations between clusters.

Standard t-SNE reveals the clustering structure at a single level of resolution.
While the perplexity parameter can be used to control the trade-off between
local and global structure, this can be time-consuming, and small, well-defined
clusters can be missed. Alternatively, \citet{kobak2019heavy} suggest that
varying the degrees of freedom in the $t$-distribution can be used to explore
the clustering structure at different levels of resolution. Modifying
Eqn.~(\ref{eq:cauchy_kernel}) to $q_{ij} \propto \left ( 1 + || \mathbf{y}_i -
\mathbf{y}_j ||^2 / \alpha \right )^{-\alpha}$ allows us to use heavier-tailed
distributions to model distances in the embedding space, which acts to highlight
small subgoups in the resulting visualization.

@@

We introduce \opentsne, a comprehensive Python library that implements t-SNE and
its numerous recently proposed extensions, including the implementation of
efficient approximation schemes allowing the embedding of millions of data
points~\citep{van2014accelerating,linderman2019fast}, the addition of new
data points into existing embeddings~\citep{policar2019embedding}, various
initialization improvements~\citep{kobak2019umap},
extensions~\citep{kobak2019heavy,kobak2019art}, and improved parameter
defaults~\citep{belkina2019automated}. Combined, these improvements result in
the faster generation of more informative visualizations.

@@

# Use the generated embeddings to add new data points
new_embedding = tsne_extensions.transform(x_new)
\end{CodeInput}
\end{CodeChunk}
The code snippet considers a \pkg{numpy} array or \pkg{scipy} sparse
matrix \texttt{x} and creates three \texttt{TSNEEmbedding} objects. These
represent fully-optimized t-SNE embeddings and are internally stored as \pkg{numpy}
arrays. These can be manipulated and visualized using standard Python tools. We
then select one of these embeddings and use it to embed new data points from
\texttt{x\_new} into the embedding space, while keeping the reference embedding
fixed. Additionally, \opentsne\ provides a more advanced interface that enables
the user to change every aspect of the t-SNE optimization procedure. This
interface is described in detail in the accompanying documentation.

% \section{Existing t-SNE Implementations}

% \opentsne\ is not the only available open-source implementation of the t-SNE 
% algorithm, however, it provides the most user-friendly experience and is the most
% feature-complete. A detailed comparison of several t-SNE implementations is
% shown in Table~\ref{tab:features}.

% \pkg{scikit-learn} is one of the most widely-used machine learning toolkits
% in the Python data science ecosystem, and provides an implementation of the
% t-SNE algorithm~\citep{pedregosa2011scikit}. However, the implementation from
% \pkg{scikit-learn} utilizes an outdated approximation scheme and is
% prohibitively slow for even moderately-sized data sets containing tens of
% thousands of samples. \opentsne\ implements all available approximation schemes,
% and defaults to the FIt-SNE approximation algorithm, which allows it to scale to
% data sets containing up to millions of data points on standard, consumer-grade,
% laptop computers. Performance benchmarks are show in Fig.~\ref{fig:benchmarks}.
% \begin{figure*}[ht!]
%   \includegraphics[width=\textwidth]{benchmarks}
%   \caption{\label{fig:benchmarks}
%   We benchmark \opentsne\ (v0.6.0) against three popular open-source
%   implementations from \pkg{scikit-learn}~\citep{pedregosa2011scikit}
%   (v0.23.1), \pkg{MulticoreTSNE}~\citep{Ulyanov2016} (v0.1), and
%   \pkg{FIt-SNE}~\citep{linderman2019fast} (v1.1.0). Experiments were run on a
%   consumer-grade Intel Core i7-7700HQ processor found in laptop computers, and
%   on a server-grade Intel Xeon E5-2650. To generate benchmark data sets of
%   different sizes, we subsampled data from the 10X Genomics 1.3 million mouse
%   brain data set five times, resulting in five different data sets for each
%   size. In total, we run each implementation on 30 different data sets. Notice
%   that \opentsne\ scales similarly to \pkg{FIt-SNE}, as they both use the
%   same interpolation-based approximation scheme, while \pkg{scikit-learn} and
%   \pkg{MulticoreTSNE} utilize the Barnes-Hut approximation.
% }
% \end{figure*}

% Pure C++ implementations, such as \pkg{BHTSNE}~\citep{van2014accelerating},
% \pkg{MulticoreTSNE}~\citep{Ulyanov2016} and
% \pkg{FIt-SNE}~\citep{linderman2019fast} are fast, but require users to
% compile and install the packages themselves. This is especially problematic for
% users of the Windows operating system, where the C/C++ compiler is not bundled
% with the operating system, making their correct configuration and compilation
% cumbersome. Additionally, these packages were not built with the Python ecosystem in
% mind, and cannot be directly used as dependencies in other Python packages.
% \opentsne\ provides several, well-documented installation options and can
% readily be used as a dependency in larger projects. We provide precompiled
% binaries through the Python Package Index (\pkg{PyPI}) or \pkg{conda} from
% the \pkg{conda-forge} channel, allowing \opentsne\ to work out-of-the-box all
% major platforms.

% Another concern is the adoption of recent theoretical improvements to the t-SNE
% algorithm. Some implementations such as \pkg{BHTSNE} and \pkg{MulticoreTSNE}
% appear to have been discontinued, as there has been little to no activity on
% their respective GitHub repositories in recent years. \pkg{scikit-learn} has a
% very active community, however, adoption of recent developments such as improved
% initialization schemes, better default parameter settings, and more flexible
% affinity kernels has been slow. Among the existing implementations,
% \pkg{FIt-SNE} includes most of the latest features but lacks crucial ones e.g.
% the ability to add new data points into existing embeddings, or flexible
% optimization strategies.

% Finally, with the exception of \pkg{scikit-learn}, the existing packages offer
% little to no user documentation, usage examples, or API reference. Additionally,
% these packages do not follow standard software development best practices, such
% as utilizing continuous integration checks or following standard style guides,
% making their maintenance difficult and unreliable.

%% -- Illustrations ------------------------------------------------------------

%% - Virtually all JSS manuscripts list source code along with the generated
%%   output. The style files provide dedicated environments for this.
%% - In R, the environments {Sinput} and {Soutput} - as produced by Sweave() or
%%   or knitr using the render_sweave() hook - are used (without the need to
%%   load Sweave.sty).
%% - Equivalently, {CodeInput} and {CodeOutput} can be used.
%% - The code input should use "the usual" command prompt in the respective
%%   software system.
%% - For R code, the prompt "R> " should be used with "+  " as the
%%   continuation prompt.
%% - Comments within the code chunks should be avoided - these should be made
%%   within the regular LaTeX text.

%\section{Illustrations} \label{sec:illustrations}

%% -- Summary/conclusions/discussion -------------------------------------------

@@

Dimensionality reduction techniques implicitly assume that high-dimensional data
lies on a lower-dimensional manifold, which can accurately be captured by a
small number of dimensions. However, there is no evidence that every data set
can accurately be described using only two dimensions, and any such embedding
will inevitably lead to a loss of information. Thus, it is beneficial to examine
multiple embeddings, each of which provides a different perspective on topology
and other data characteristics.

\begin{figure*}[htbp]
  \center
  \includegraphics[width=0.75\textwidth]{tasic2018}
  \caption{\label{fig:tasic}
  We use \opentsne\ to create four different
  visualizations of the \citet{tasic2018shared} data,
  each providing a different perspective into the topology of the data.
  The data set contains 21,874 single-cells originating from the mouse
  neocortex. Cluster annotations and colors are taken from the original
  publication. Warm colors correspond to excitatory neurons, cool colors
  correspond to inhibitory neurons, and gray/brown colors correspond to
  non-neuronal cells. Standard t-SNE \textbf{(a)} emphasizes local
  structure while increasing perplexity \textbf{(b)} results in a more
  meaningful layout of the clusters. We can also combine the two
  perplexities by using a multiscale kernel affinity model \textbf{(c)}
  and obtain a trade-off between global and local structure.
  Alternatively, we can inspect more fine-grained structure and reveal
  smaller clusters by using a more heavy-tailed kernel \textbf{(d)}.}
\end{figure*}

@@

\section{Conclusion}
We introduce \opentsne, the most feature-complete, open-source, Python
implementation of the widely popular t-SNE visualization algorithm. The library
incorporates the latest developments to the t-SNE algorithm and makes them
easily accessible through a familiar API. \opentsne\ implements efficient
approximation schemes, making it suitable for the visualization of large data
sets, containing up to millions of data points. Finally, \opentsne\ is easy to
install and comes with precompiled binaries available for all major platforms.

>>>>

%% - \Abstract{} almost as usual
\Abstract{
One of the most popular techniques for constructing visualizations of large, high-dimensional data sets is $t$-distributed stochastic neighbor embedding (t-SNE). Recently, several extensions have been proposed to address scalability issues and the quality of the resulting visualizations. We introduce \opentsne, a modular Python library that implements the core t-SNE algorithm and its many extensions. The library is fast and can treat data sets containing millions of data points. We distribute \opentsne\ under the BSD-3-Clause License. Its source code is available at \url{https://github.com/pavlin-policar/openTSNE}.
}

@@

The ever-growing volumes of high-dimensional data sets in machine learning call for efficient dimensionality reduction techniques and their implementations, capable of producing informative data visualizations. Popular approaches include principal component analysis, multidimensional scaling, $t$-distributed stochastic neighbor embedding (t-SNE)~\citep{maaten2008visualizing}, and uniform manifold approximation and projections (UMAP)~\citep{2018arXivUMAP}. Among these, t-SNE has received much attention as it can address high volumes of data and reveal the underlying data structure. For instance, t-SNE is widely used in the bioinformatics community in areas such as single-cell transcriptomics~\citep{macosko2015highly,cao2019single,tasic2018shared}, human genetics~\citep{hirata2019genetic}, metagenomic assembly~\citep{beaulaurier2018metagenomic}, the spatial organization of microbial communities~\citep{sheth2019spatial}, and metabolomics~\citep{tkachev2019differences}. To visually explore the data structure, the objects of interest, such as diseased and healthy tissues or single cells, are profiled through thousands of features that include, for example, expression of genes or concentration of metabolites. The task of t-SNE is to embed the objects of interest within a low, preferably two-dimensional embedding. Reports on single-cell gene expression data, our running example, often start with an overview of the cell landscape and use t-SNE to embed high-dimensional expression profiles into a two-dimensional space. Figs.~\ref{fig:macosko}.a and \ref{fig:macosko}.b show two such embeddings.

\begin{figure*}[ht]
  \includegraphics[width=\textwidth]{macosko2015}
  \caption{\label{fig:macosko}
We use \opentsne\ to generate three t-SNE embeddings and demonstrate recent theoretical advances. The data in \textbf{(a)} and \textbf{(b)} includes 44,808 mouse retinal cells (dots in the plots) that are described with high-dimensional gene-expression profiles from \citet{macosko2015highly}. The data in \textbf{(c)} additionally contains 27,499 expression profiles from mouse retinal cells from \citet{shekhar2016comprehensive}. \textbf{(a)} We construct t-SNE embedding following the parameter choices from the original publication by \citet{maaten2008visualizing}. The visualization shows no preservation of the global organization of clusters, resulting from random initialization and an affinity model focused on preserving local neighborhoods. \textbf{(b)} A modern t-SNE embedding, utilizing the latest theoretical advances and practical recommendations constructed using a multi-scale affinity model, preserving both short-range and long-range interactions between data points and initialized so that the global layout is as meaningful as possible. Unlike in \textbf{(a)}, the green and blue clusters representing different sub-types of amacrine and bipolar cells are now localized to the same regions of the space, indicating a higher level of similarity than to other cell types. The embedding in \textbf{(c)} shows how existing t-SNE reference atlases can be used to place new samples into existing embeddings. The positions of new data points correspond to cell types from the reference atlas.
}
\end{figure*}

Despite its utility, t-SNE has often been criticized for its limited scalability, lack of global organization -- t-SNE identifies well-defined clusters that may be arbitrarily scattered throughout the embedding, and the absence of theoretically-founded methods to map new data into existing embeddings~\citep{ding2018interpretable,becht2019dimensionality}. Most of these shortcomings have recently been addressed. \citet{linderman2019fast} developed FIt-SNE, an efficient approximation scheme that massively improves the scalability of t-SNE, achieving linear time complexity in the number of samples. \citet{kobak2019art} proposed several techniques to improve global cluster coherence, including estimating similarities with a mixture of Gaussian kernels. In our previous work, we introduced a principled approach for embedding new samples into existing visualizations~\citep{policar2019embedding}.

However, easy access to efficient implementations almost universally correlates with the widespread adoption of novel computational techniques. Despite the many theoretical advancements, the most popular t-SNE libraries have been slow to incorporate them into their implementations. In this paper, we present \opentsne, our open-source Python implementation of t-SNE, and the recently proposed extensions to t-SNE that \opentsne\ implements. \opentsne\ is easy to install and includes precompiled binaries available through popular Python package managers. It provides a familiar API, making it suitable as a drop-in replacement for existing t-SNE implementations.

@@

In standard t-SNE, distances between data points are converted to similarities through the use of Gaussian kernels. The perplexity of these kernels governs the sizes of the local neighborhoods which are to be preserved. One easy way to improve the global consistency of resulting embeddings is to increase the sizes of these neighborhoods. This can be achieved by increasing the perplexity parameter. However, this often comes at the price of local structure. \citet{kobak2019art} propose to use mixtures of Gaussians instead to preserve better short-range and long-range distances and, thus, achieve a better trade-off between local and global structure.

Embeddings produced by standard t-SNE often use all available space and separate clusters by only thin boundaries. When working with large data sets, this often obscures the global relationships between clusters as all neighboring groups appear equidistant from one another.  Other dimensionality reduction methods, such as UMAP and ForceAtlas2, produce embeddings where clusters appear more compact, and the white space separating the clusters may be interpreted as a loose measure of distance. Recently, \citet{bohm2020unifying} showed that the exaggeration factor $\rho$ could be used to produce layouts more similar to UMAP and ForceAtlas2. By incorporating exaggeration into later phases of the optimization, t-SNE introduces white space between clusters, which better reflects the global relations between clusters.

Standard t-SNE reveals the clustering structure at a single level of resolution. While the perplexity parameter can be used to control the trade-off between local and global structure, this can be time-consuming, and small, well-defined clusters can be missed. Alternatively, \citet{kobak2019heavy} suggest that varying the degrees of freedom in the $t$-distribution can be used to explore the clustering structure at different levels of resolution. Modifying Eqn.~(\ref{eq:cauchy_kernel}) to $q_{ij} \propto \left ( 1 + || \mathbf{y}_i - \mathbf{y}_j ||^2 / \alpha \right )^{-\alpha}$ allows us to use heavier-tailed distributions to model distances in the embedding space, which acts to highlight small subgroups in the resulting visualization.

@@

We introduce \opentsne, a comprehensive Python library that implements t-SNE and its numerous recently proposed extensions, including the implementation of efficient approximation schemes allowing the embedding of millions of data points~\citep{van2014accelerating,linderman2019fast}, the addition of new data points into existing embeddings~\citep{policar2019embedding}, various initialization improvements~\citep{kobak2019umap}, extensions~\citep{kobak2019heavy,kobak2019art}, and improved parameter defaults~\citep{belkina2019automated}. When combined, these improvements result in the faster generation of more informative visualizations.

@@

# Use the generated embeddings to add new data points
new_embedding = tsne_extensions.transform(x_new)
\end{CodeInput}
\end{CodeChunk}
The code snippet considers a \pkg{numpy} array or \pkg{scipy} sparse matrix \texttt{x} and creates three \texttt{TSNEEmbedding} objects. These represent fully-optimized t-SNE embeddings, and \opentsne\ internally stores them as \pkg{numpy} arrays. These can be manipulated and visualized using standard Python tools. We then select one of these embeddings and use it to embed new data points from \texttt{x\_new} into the embedding space while keeping the reference embedding fixed. Additionally, \opentsne\ provides a more advanced interface that enables the user to change every aspect of the t-SNE optimization procedure. We describe this interface in detail in the \opentsne's on-line documentation.

@@

Dimensionality reduction techniques implicitly assume that high-dimensional data lies on a lower-dimensional manifold, which a small number of dimensions can accurately capture. However, there is no evidence that every data set can accurately be described using only two dimensions, and any such embedding will inevitably lead to a loss of information. Thus, it is beneficial to examine multiple embeddings, each of which provides a different perspective on topology and other data characteristics.

\begin{figure*}[htbp]
  \center
  \includegraphics[width=0.75\textwidth]{tasic2018}
  \caption{\label{fig:tasic}
  We use \opentsne\ to create four different
  visualizations of the \citet{tasic2018shared} data,
  each providing a different perspective into the topology of the data.
  The data set contains 21,874 single-cells originating from the mouse
  neocortex. Cluster annotations and colors are taken from the original
  publication. Warm colors correspond to excitatory neurons, cool colors
  correspond to inhibitory neurons, and gray/brown colors correspond to
  non-neuronal cells. Standard t-SNE \textbf{(a)} emphasizes local
  structure while increasing perplexity \textbf{(b)} results in a more
  meaningful layout of the clusters. We can also combine the two
  perplexities by using a multi-scale kernel affinity model \textbf{(c)}
  and obtain a trade-off between global and local structure.
  Alternatively, we can inspect more fine-grained structure and reveal
  smaller clusters by using a more heavy-tailed kernel \textbf{(d)}.}
\end{figure*}

@@

\section{Conclusion}
We introduce \opentsne, the most feature-complete, open-source, Python implementation of the widely popular t-SNE visualization algorithm. The library incorporates the latest developments to the t-SNE algorithm and makes them easily accessible through a familiar API. \opentsne\ implements efficient approximation schemes, making it suitable for the visualization of large data sets containing up to millions of data points. Finally, \opentsne\ is easy to install and comes with precompiled binaries available for all major platforms.
